---
title: "U.S. Severe Weather Events Economic and Health Impacts"
output: html_document
---

#Summary
Using data from the NOAA Storm Database, I've determined the top 10 types of storm events measured by their impact to both human health, based on the number of reported injuries and fatalities, and the economic value of properties, goods, and resources. The key findings are that the top 5 event types impacting human health are [] and the top 5 event types with harmful economic impact are []. 

#Data Processing


Load libraries used in the analysis.

```{r}
library(dplyr)
library(stringdist)
```


Download the source data file from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database and load it into a data frame.

```{r}
url <- c("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2")

# Commenting out download for now
#download.file(url, destfile = "StormData.csv.bz2")

# Needs to remove path to Desktop for final submission
df <- read.csv(bzfile("/Users/markkaufman/Desktop/StormData.csv.bz2"))


# to backup when I screw things up
df_backup <- df
```

Create Year Column.

```{r}
df$year <- format(strptime(df$BGN_DATE, "%m/%d/%Y %H:%M:%S"),"%Y")
```

Create Combined Fatalities and Injuries Column.

```{r}
df$combined <- df$INJURIES + df$FATALITIES
```

Simple clean up values of EVTYPES, PROPDMGEXP, and CROPDMGEXP.

```{r}
# Capitalize all EVTYPE values
df$EVTYPE <- as.character(df$EVTYPE)
df$EVTYPE <- toupper(df$EVTYPE)

# Capitalize all PROPDMGEXP and CROPDMGEXP values
df$PROPDMGEXP <- as.character(df$PROPDMGEXP)
df$PROPDMGEXP <- toupper(df$PROPDMGEXP)
df$CROPDMGEXP <- as.character(df$CROPDMGEXP)
df$CROPDMGEXP <- toupper(df$CROPDMGEXP)

```


Compute the economic damage (total_damage) for each event. First need to translate some of the values in Property Damage (PROPDMGEXP) and Crop Damange (CROPDMGEXP) columns to numbers that can be used in the calculations. 

```{r}

df$PROPDMGEXP[df$PROPDMGEXP == "" | df$PROPDMGEXP == "-" | df$PROPDMGEXP == "?" | df$PROPDMGEXP == "+"] <- 0
df$PROPDMGEXP[df$PROPDMGEXP == "H"] <- 2
df$PROPDMGEXP[df$PROPDMGEXP == "K"] <- 3
df$PROPDMGEXP[df$PROPDMGEXP == "M"] <- 6

df$CROPDMGEXP[df$CROPDMGEXP == "" | df$CROPDMGEXP == "-" | df$CROPDMGEXP == "?" | df$CROPDMGEXP == "+"] <- 0
df$CROPDMGEXP[df$CROPDMGEXP == "H"] <- 2
df$CROPDMGEXP[df$CROPDMGEXP == "K"] <- 3
df$CROPDMGEXP[df$CROPDMGEXP == "M"] <- 6

df$PROPDMGEXP <- as.numeric(df$PROPDMGEXP)
df$CROPDMGEXP <- as.numeric(df$CROPDMGEXP)
df$total_damage <- df$PROPDMG * 10^df$PROPDMGEXP + df$CROPDMG * 10^df$CROPDMGEXP

```


Create table of official storm list EVTYPES.
```{r}

events = c("ASTRONOMICAL LOW TIDE", "AVALANCHE", "BLIZZARD", "COASTAL FLOOD", "COLD/WIND CHILL", "DEBRIS FLOW", "DENSE FOG", "DENSE SMOKE", "DROUGHT", "DUST DEVIL", "DUST STORM", "EXCESSIVE HEAT", "EXTREME COLD/WIND CHILL", "FLASH FLOOD", "FLOOD", "FROST/FREEZE", "FUNNEL CLOUD", "FREEZING FOG", "HAIL", "HEAT", "HEAVY RAIN", "HEAVY SNOW", "HIGH SURF", "HIGH WIND", "HURRICANE (TYPHOON)", "ICE STORM", "LAKE-EFFECT SNOW", "LAKESHORE FLOOD", "LIGHTNING", "MARINE HAIL", "MARINE HIGH WIND", "MARINE STRONG WIND", "MARINE THUNDERSTORM WIND", "RIP CURRENT", "SEICHE", "SLEET", "STORM SURGE/TIDE", "STRONG WIND", "THUNDERSTORM WIND", "TORNADO", "TROPICAL DEPRESSION", "TROPICAL STORM", "TSUNAMI", "VOLCANIC ASH", "WATERSPOUT", "WILDFIRE", "WINTER STORM", "WINTER WEATHER")

# Maybe not supposed to be a data frame? PROBABLY DELETE THIS
officialstormlist <- data.frame(EVTYPES = c("ASTRONOMICAL LOW TIDE", "AVALANCHE", "BLIZZARD", "COASTAL FLOOD", "COLD/WIND CHILL", "DEBRIS FLOW", "DENSE FOG", "DENSE SMOKE", "DROUGHT", "DUST DEVIL", "DUST STORM", "EXCESSIVE HEAT", "EXTREME COLD/WIND CHILL", "FLASH FLOOD", "FLOOD", "FROST/FREEZE", "FUNNEL CLOUD", "FREEZING FOG", "HAIL", "HEAT", "HEAVY RAIN", "HEAVY SNOW", "HIGH SURF", "HIGH WIND", "HURRICANE (TYPHOON)", "ICE STORM", "LAKE-EFFECT SNOW", "LAKESHORE FLOOD", "LIGHTNING", "MARINE HAIL", "MARINE HIGH WIND", "MARINE STRONG WIND", "MARINE THUNDERSTORM WIND", "RIP CURRENT", "SEICHE", "SLEET", "STORM SURGE/TIDE", "STRONG WIND", "THUNDERSTORM WIND", "TORNADO", "TROPICAL DEPRESSION", "TROPICAL STORM", "TSUNAMI", "VOLCANIC ASH", "WATERSPOUT", "WILDFIRE", "WINTER STORM", "WINTER WEATHER"))

```


Some manual clean-up of EVTYPES 

```{r}
df$EVTYPE[df$EVTYPE == "EXTREME COLD"] <- "EXTREME COLD/WIND CHILL"
df$EVTYPE[df$EVTYPE == "TSTM WIND"] <- "THUNDERSTORM WIND"
df$EVTYPE[df$EVTYPE == "TYPHOON"] <- "TROPICAL STORM"
```


Clean up the EVTYPEs, which are a mess. 

```{r}
# Create copy of evtype with all 9000+ obs
evtype <- (df$EVTYPE)

## match the dictionary and event types
match2 <- amatch(evtype, events, maxDist=2)

## Create frequency table for matched obs. 
#library(plyr)
match1 <- (events[match2])
#matches_recorded <- count(match1)


## This results in 47 vars, including one `NA` var
# Perhaps 2 official event type names were not used, hence 47 obs

## Check matching success
# obs 47 is the NA col, so (sum of all vars except NA) / total sum

#> sum(matches_recorded$freq[-47])/sum(matches_recorded$freq) 
#0.7298362 - CONTROL
#0.9744297 - my match rate 
```


Then swap in the official list values for EVTYPE

```{r}
df$EVTYPE <- match1
```


MORE possible clean-up: Strip leading tabs or spaces that are at the beginning of names. example REFNUM 448421, which begins with a tab
```{r}

```


Remove data that skews the analysis. Although the NOAA started reporting Tornadoes in 1950, the official 48 event types did not start reporting until 1996,

```{r}
df_econ <- filter(df, year > 1995)
```


Determine the cost and number of combined fatalities and injuries per event type. Also compute average cost per event and average fatalities and injuries per event.

```{r}
# If needed to unload plyr
# detach("package:plyr", unload=TRUE) 

by_EVTYPE <- group_by(df_econ, EVTYPE)
by_EVTYPE_total_cost <- summarise(by_EVTYPE, total_cost = sum(total_damage), INJURIES = sum(INJURIES), FATALITIES = sum(FATALITIES), injuries_and_death = sum(combined), count = n())

# Compute Average cost per event
by_EVTYPE_total_cost$avg_cost_per_event <- as.integer(by_EVTYPE_total_cost$total_cost / by_EVTYPE_total_cost$count)


# Compute Average fatalities and injuries per event
by_EVTYPE_total_cost$avg_injuries_and_death_per_event <- as.integer(by_EVTYPE_total_cost$injuries_and_death / by_EVTYPE_total_cost$count)

```

plot the top 10 events by total cost

```{r}
library(ggplot2)
library(gridExtra)

top10econ <- arrange(by_EVTYPE_total_cost, desc(total_cost))[1:10,]

top10econ$total_cost <- top10econ$total_cost/10^9

plot1 <- ggplot(top10econ, aes(x = EVTYPE, y = total_cost))
plot1 <- plot1 + geom_bar(stat = "identity") + labs(title = "Total Cost", x = "Event Type", y = "Economic Impact (Billions U.S. Dollars)") + theme(axis.text.x = element_text(angle = 90, hjust = 1))



```

plot the number of events for the top 10 events by total cost

```{r}
plot2 <- ggplot(top10econ, aes(x = EVTYPE, y = count))
plot2 <- plot2 + geom_bar(stat = "identity") + labs(title = "Number of Events", x = "Event Type", y = "Number of recorded events (1996-2011)") + theme(axis.text.x = element_text(angle = 90, hjust = 1))

```


```{r}

require(gridExtra)
grid.arrange(plot1, plot2, ncol=2)

```



plot the avg economic impact per event for the top 10 events by total cost
```{r}

plot3 <- ggplot(top10econ, aes(x = EVTYPE, y = avg_cost_per_event))
plot3 <- plot3 + geom_bar(stat = "identity") + labs(title = "Average Cost", x = "Event Type", y = "Average Cost Per Event (U.S. Dollars)") + theme(axis.text.x = element_text(angle = 90, hjust = 1))

```



```{r}

require(gridExtra)
grid.arrange(plot1, plot2, plot3, ncol=2, nrow=2, top = "Top Economically Harmful U.S. Storm Event Types (1996-2011)")

```



plot the top 10 events based on total injuries AND fatalities 

```{r}
top10injuries_and_death <- arrange(by_EVTYPE_total_cost, desc(injuries_and_death))[1:10,]

c <- ggplot(top10injuries_and_death, aes(x = EVTYPE, y = injuries_and_death))
c + geom_bar(stat = "identity") + labs(title = "Top 10 U.S. Storm Event Types Causing Most Human Health Harm (1996-2011)", x = "Event Type", y = "Number of Fatalities and Injuries") + theme(axis.text.x = element_text(angle = 90, hjust = 1))

```



plot the top 10 events based on total injuries 

```{r}
top10injuries <- arrange(by_EVTYPE_total_cost, desc(INJURIES))[1:10,]

plotb1 <- ggplot(top10injuries, aes(x = EVTYPE, y = INJURIES))
plotb1 <- plotb1 + geom_bar(stat = "identity") + labs(title = "Number of Injuries", x = "Event Type", y = "Number of Injuries") + theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

plot number of events  - injuries

```{r}

plotb2 <- ggplot(top10injuries, aes(x = EVTYPE, y = count))
plotb2 <- plotb2 + geom_bar(stat = "identity") + labs(title = "Number of Recorded Events", x = "Event Type", y = "Number of Events Recorded") + theme(axis.text.x = element_text(angle = 90, hjust = 1))

```


```{r}

require(gridExtra)
grid.arrange(plotb1, plotb2, ncol=2, top = "Top U.S. Storm Event Types Causing Human Injury (1996-2011)")

```




plot the top 10 events based on total fatalities 
```{r}
top10fatalities <- arrange(by_EVTYPE_total_cost, desc(FATALITIES))[1:10,]

plotc1 <- ggplot(top10fatalities, aes(x = EVTYPE, y = FATALITIES))
plotc1 <- plotc1 + geom_bar(stat = "identity") + labs(title = "Fatalities", x = "Event Type", y = "Number of Fatalities") + theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

plot number of events - fatalities

```{r}

plotc2 <- ggplot(top10fatalities, aes(x = EVTYPE, y = count))
plotc2 <- plotc2 + geom_bar(stat = "identity") + labs(title = "Number of Recorded Events", x = "Event Type", y = "Number of Events Recorded") + theme(axis.text.x = element_text(angle = 90, hjust = 1))

```


```{r}

require(gridExtra)
grid.arrange(plotc1, plotc2, ncol=2, top = "Top U.S. Storm Event Types Causing Fatalities (1996-2011)")

```


